### 2021-07-07 rad-icfp-2021/generalization.tex
.

\section{Beyond the Source language}
\label{sec:generalization}

\subsection{Lifting the restriction on reduce}
\label{sub:Lifting the restriction on reduce}

Assume we allow $\Gamma$, !x!: $\reals$, !y!: $\reals$ $\vdash$ !e!: $\reals$ to be the function argument in
!reduce (x,y.e) v A!. We need to add a term depending on $\grad_{\{xi\}}e$ to !yi! in the continuation.
Writing the derivative becomes quite cumbersome, the notation is more compact using arrays of tuples. 
Similarly to he monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$.

We can now compute the reverse-mode transformation as follows.
\begin{center}
\begin{tabular}{r c l}
$\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e1) e$_2$ e$_3$!) 
&=& $\ldots$ as before up until !A_$3$! \\
&& !let B$_0$ = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A$_0$ A) in!\\
&& !in let B$_1$ = scanr $1_\Gamma$ $\widehat{\times}$ B$_0$ in! \\
&& !let B$_2$ = reduce $\widehat{+}$ $0_\Gamma$ B$_1$ in! \\
&& !<reduce (x,y.e1) y$_1$ A, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}$B$_2$,0,map2 (a,b. a*b*z) A$_2$ A$_3$)!
\end{tabular}
\end{center}

\subsection{More array operators}
\label{sub:Adding more array operators}

@@

Finally, the former point implies we need a few more array manipulations.
These can be elegantly dealt with by changing the semantics of !scanl! and !scanr!. 
Let's assume that they now return a pair of an array of size $n$ 
of the intermediate computations and the final result.
The reverse derivative of fold left is then

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$(!foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
            !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
            && !let A,Y$_2$ = $\directD{\rho}{\Gamma,v}{Y_1}$(e$_3$) in! \\
            && !let A$_0$,r$_1$ = (scanl (x,y.e$_1$) v A) in! \\
            && !let A$_1$ = map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x,b/y]) A$_0$ A! \\
            && !let A$_2$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_1$[a/x,b/y]) A$_0$ A! \\
            && !let r$_2$, A$_3$ = scanr * 1 A$_1$! \\
            && !<r$_1$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
            && !let y$_1$,B = r$_2$*z ,map2 (x,y. x*y*z) A$_2$ A$_3$ in! \\
            && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
\end{tabular}
\end{center}

Another useful example is !map!, which is very similar to !map2!.

\begin{center}
    \begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
            !let A,Y$_{1}$ = $\directD{\rho}{\Gamma}{Y}$(e$_{2}$) in! \\
            && !let G=$\grad_{\Gamma}$ e$_{1}$ in!\\
            && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)$\widehat{+}$(G*(reduce + 0 Z)),!\\
            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\
    \end{tabular}
\end{center}

\subsection{Non-smooth scalar operators} % (fold)
\label{sub:Adding more scalar operators}

\subsection{Conditionals} % (fold)
\label{sub:Adding conditionals}

Adding conditionals to the source language can be done easily, for instance via defining

@@

A slightly better option would be to define

@@

It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
but there is still a non-linear usage of the continuation variable !Y!.
One solution, if we know we want to compute the whole gradient of the expression, is to define instead

@@

\subsection{General arrays} % (fold)
\label{sub:Adding general arrays}

@@

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
For !map2!, we need a generalization of $*:\reals\to\reals$. 
If !e1: A!, then $\grad_{\{x\}}$!e1: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
$\widetilde{*}$ corresponds to matrix-matrix multiplication.
Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!.

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
    \end{tabular}
    \end{center}

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
    && !let A$_{1}$ = shift1L (map2! \\ 
    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
    && !let A$_{2}$ = map2! \\
    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\
\end{tabular}
\end{center}

>>>>

\section{Beyond the Source language}
\label{sec:generalization}
In this section, we generalize our language by relaxing the imposed restrictions.
First, we allow free variables for the function argument of !reduce! in Section~\ref{sub:lift_reduce}. 
Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
Finally, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}.

\subsection{Lifting the restriction on reduce}
\label{sub:lift_reduce}

Assume we allow $\Gamma$, !x!: $\reals$, !y!: $\reals$ $\vdash$ !e!: $\reals$ to be the function argument in
!reduce (x,y.e) v A!. We need to add a term depending on $\grad_{\{xi\}}e$ to !yi! in the continuation.
As writing the derivative becomes quite cumbersome, we use a more compact notation using arrays of tuples. 
Similarly to the monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$. The modified reverse-mode transformation is shown in Figure~\ref{fig:lift_reduce}.

We can now compute the reverse-mode transformation as follows.

\begin{figure}

\begin{small}
\begin{tabular}{|r c l|}
\hline
$\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e1) e$_2$ e$_3$!) 
&=& $\ldots$ same as Figure~\ref{fig:direct_diff_macro} up until !A_$3$! \\
&& !let B$_0$=map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A$_0$ A) in!\\
&& !let B$_1$=scanr $1_\Gamma$ $\widehat{\times}$ B$_0$ in! \\
&& !let B$_2$=reduce $\widehat{+}$ $0_\Gamma$ B$_1$ in! \\
&& !<reduce (x,y.e1) y$_1$ A, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}$B$_2$,0,map2 (a,b. a*b*z) A$_2$ A$_3$)! \\ \hline
\end{tabular}
\end{small}
\caption{The reverse-mode AD transformation of \codekw{reduce} without restrictions on its function argument.}
\label{fig:lift_reduce}
\end{figure}

\subsection{More array operators}
\label{sub:lift_more_arr_op}

@@

Finally, the former point implies we need a few more array manipulations.
These can be elegantly dealt with by changing the semantics of !scanl! and !scanr!. 
Let's assume that they now return a pair of an array of size $n$ 
of the intermediate computations and the final result.
The reverse derivative of !foldl! is then as shown in Figure~\ref{fig:foldl_map_trans}.
In this figure, we have shown the translation of the !map! operator as well, which is very similar to !map2!.

\begin{figure}
\begin{center}
\begin{tabular}{|r c l|}
\hline
    $\directD{\rho}{\Gamma}{Y}$(!foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
            !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
            && !let A,Y$_2$ = $\directD{\rho}{\Gamma,v}{Y_1}$(e$_3$) in! \\
            && !let A$_0$,r$_1$ = (scanl (x,y.e$_1$) v A) in! \\
            && !let A$_1$ = map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x,b/y]) A$_0$ A! \\
            && !let A$_2$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_1$[a/x,b/y]) A$_0$ A! \\
            && !let r$_2$, A$_3$ = scanr * 1 A$_1$! \\
            && !<r$_1$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
            && !let y$_1$,B = r$_2$*z ,map2 (x,y. x*y*z) A$_2$ A$_3$ in! \\
            && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
    $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
            !let A,Y$_{1}$ = $\directD{\rho}{\Gamma}{Y}$(e$_{2}$) in! \\
            && !let G=$\grad_{\Gamma}$ e$_{1}$ in!\\
            && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)$\widehat{+}$(G*(reduce + 0 Z)),!\\
            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\\hline
\end{tabular}
\end{center}
\caption{The reverse-mode AD transformation of \codekw{foldl} and \codekw{map} operators.}
\label{fig:foldl_map_trans}
\end{figure}

\subsection{Conditionals} % (fold)
\label{sub:lift_conditional}

\smartpara{Non-Solution 1}
Adding conditionals to the source language can be done easily, for instance via defining

@@

\smartpara{Non-Solution 2}
A slightly better option would be to define

@@

It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
but there is still a non-linear usage of the continuation variable !Y!.

\smartpara{Solution}
If we know we want to compute the whole gradient of the expression, we can define the translation as follows:

@@

\subsection{Non-smooth scalar operators} % (fold)
\label{sub:lift_non_smooth}

\subsection{General arrays} % (fold)
\label{sub:lift_gen_arr}

@@

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
For !map2!, we need a generalization of $*:\reals\to\reals$. 
If !e1: A!, then $\grad_{\{x\}}$!e1: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
$\widetilde{*}$ corresponds to matrix-matrix multiplication.
Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.

\begin{figure}
\begin{center}
\begin{tabular}{|r c l|}
\hline
    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
    && !let A$_{1}$ = shift1L (map2! \\ 
    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
    && !let A$_{2}$ = map2! \\
    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\ \hline
\end{tabular}
\end{center}
\caption{The reverse-mode AD transformation of \codekw{map2} and \codekw{reduce} for general arrays.}
\label{fig:lift_gen_arr}
\end{figure}
